### 2011-01-20 paper/approach.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

\section{Introduction}

% Motivate and abstractly describe the problem you are addressing and
% how you are addressing it. What is the problem? Why is it important?
% What is your basic approach? A short discussion of how it fits into
% % related work in the area is also desirable. Summarize the basic
% results and conclusions that you will present.

Applications of walking robots often call for the ability to walk as
quickly, efficiently, or with as little power as possible.  Often
these optimizations are done manually by an expert who designs and
tweaks a gait specifically for a given objective.  Other groups have
used learning methods to generate gaits optimized for some metric.
Approaches differ in their starting assumptions, some essentially
tweaking the parameters of a hand-tuned model \cite{chernova}, others
exploring a reasonably compact parameter space \cite{kohl}, and still
others beginning with few assumptions besides periodicity
\cite{zykov}.

We aimed to strike a middle ground between these approaches.  Our
motion generator did not rely on hand-tweaked gaits, but it did use
parameterized gaits which, by their nature, impose some assumptions on
the answers produced.  We then used machine learning to design gaits
for a quadruped robot with these models.  This paper presents a
comparison of the different learning methods implemented.  Most
methods created walks that are several times faster than the original
hand-tuned gait.  We invite readers with short attention spans to view
a video of some of our results online here:

%\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
\texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}

%The code so far consists of the robot class, optimization class,
%parameterized model, sine model, and camera feedback.

\section{Problem definition}
% Precisely define the problem you are addressing (i.e. formally specify
% the inputs and outputs). Elaborate on why this is an interesting and
% important problem.

We are testing several different learning methods to design a
parametrized gait for a quadruped robot from the Cornell Computational
Synthesis Lab.

The output each of the learning algorithms is a function of time,
$f(t)$, that outputs a vector of commanded motor positions.  This
function is generated using a parametrized motion model, described in
\secref{implement}.

The robot executes these commands and measures its change in location
using the tracking system described in \secref{implement}.  The input
to the learning algorithms is this measured displacement, which the
algorithms attempt to maximize. This displacement is measured for each
gait over a constant length run, usually 12 seconds.

A comparison and evaluation of the many different methods available
for optimizing the gait of legged robots will be useful for future
work on this challenging multidimensional control problem.

\section{Method}
\seclabel{method}

@@

\section{Related work}
% Briefly explain who else worked on related problems in the past and what
% methods they used. Explain if you are using similar methods, or if your
% approach is different and if so - how (either is ok).

Various maching learning techniques have proven to be useful in
finding control policies for a wide variety of robots. Kohl and
Stone\cite{kohl} presented a policy gradient reinforcement learning
approach for generating a fast walk on legged robots. We experimented
with this method to create a walk for our robot (Policy Gradient
Descent). Chernova and Velosa\cite{chernova} took an evolutionary
approach to this problem which we did not implement.  Zykov, Bongard,
and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
physical robot requiring no prior assumptions about the locomotion
pattern beyond the fact that it should be rhythmic.

\section{System Architecture and Implementation}
\seclabel{implement}
% Describe how you implemented your system and how you structured it. 
% This should give an overview of the system, not a detailed 
% documentation of the code. The documentation of the code is part of 
% the code you hand in. You might want to comment on high-level design 
% decisions that you made. Also explain how you obtained your
% data and any pre-processing you did to it.

@@

\section{Experimental Evaluation}

\subsection{Methodology}
% What are the criteria you are using to evaluate your method? What
% specific hypotheses does your experiment test? Describe the 
% experimental methodology that you used. What are the dependent and 
% independent variables? For projects in machine learning, what is 
% the training/test data that was used, and why is it realistic or
% interesting? Exactly what performance data did you collect and how 
% are you presenting and analyzing it? Comparisons to competing methods 
% that address the same problem or to variations of your own algorithm 
% are particularly useful.

@@

\subsection{Results}
% Present the quantitative results of your experiments. Graphical data
% presentation such as graphs and histograms are frequently better 
% than tables. Explain what are the basic differences revealed in the 
% data. If your method is stochastic or you analyze multiple datasets, 
% calculate mean and standard error for every quantitative
% performance result you show based on multiple runs.

We have done complete runs of 3 different initial parameter vectors
for random search, uniform random hill climbing, Gaussian random hill
climbing, policy gradient descent, Nelder-Mead simplex, and linear
regression.  We also evaluated the SVM regression and HyperNEAT
methods, but these methods were more experimental, and thus we do not
yet have the same volume of data for these runs.  We developed several
gaits that were about 4 times faster than the original hand-coded
gait.  Results are shown in \tabref{results_table}.

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & A       & B      & C      &  Average \\
\hline                                                               
\hline                                                               
Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
\hline                                                                 
Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
\hline                                                                 
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
\hline                                                                 
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
\hline                                                                 
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
\hline                                                                 
Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
\hline                                                                 
Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
%\hline                                                                
%SVM Regression                           & ???     & ???    & ???   &   \\
%\hline                                                                
%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
\hline
\end{tabular}
\caption{The best gaits found for each starting vector and algorithm,
  in body lengths per minute.}
\tablabel{results_table}
\end{center}
\end{table}

\item For vector A, shown in \figref{vectorA}, linear regression
  worked significantly better than the other algorithms, resulting in
  a gait (27.58 body lengths/minute) that walked over twice as fast as
  the next best gait -- uniform random hill climbing at 11.37 body
  lengths/minute -- and 5.3 times better than the previous hand-coded
  gait.

\item Vector B, depicted in \figref{vectorB}, resulted in similarly
  performing gaits with all the algorithms. The random method got
  lucky and produced the best gait (17.26 body lengths/minute) and
  uniform random hill climbing produced the worst (9.44 body
  lengths/minute). The remaining algorithms each produced a gait
  around 13 body lengths/minute.

\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
  random hill climbing each produced a gait that substantially
  outperformed the other algorithms. Simplex resulted in a gait of
  nearly 15 body lengths/minute and Gaussian random hill climbing
  produced a gait of just over 13 body lengths/minute, whereas the
  other algorithms returned gaits of less than 5 body lengths/minute.

\figvarp{vectorA}{1}{Results for runs beginning with vector A.
  Linear regression worked significantly better than the other
  algorithms, resulting in a gait (27.58 body lengths/minute) that
  walked over twice as fast as the next best gait -- uniform random
  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
  than the previous hand-coded gait.}{}

\figvarp{vectorB}{1}{Results for run starting with vector B.  All
  algorithms performed similarly.  The random method actually got
  lucky this time and produced the best gait (17.26 body
  lengths/minute) and uniform random hill climbing produced the worst
  (9.44 body lengths/minute). The remaining algorithms each produced a
  gait around 13 body lengths/minute.}{}

\figvarp{vectorC}{1}{Results for runs starting with vector C.
  Simplex and Gaussian random hill climbing each produced a gait that
  substantially outperformed the other algorithms, with around 15 body
  lengths/minute and 13 body lengths/minute, respectively, whereas the
  other algorithms returned gaits of less than 5 body
  lengths/minute.}{}

Based on the three trials, as shown in \figref{std_error}, linear
regression worked the best, followed by simplex. Gaussian random hill
climbing performed about as well as random, but it is unclear whether
Gaussian would continue to improve and become much better than random
search if allowed to run for more iterations. Uniform random hill
climbing and policy gradient descent performed similarly to each other
on average, but differed greatly on individual runs.

\figvarp{std_error}{1}{Average results for each method starting with
  parameter vectors A, B, and C.  Error bars are plotted showing the
  standard error.  Linear regression outperformed other methods,
  followed by simplex. Gaussian random hill climbing performed about
  as well as random, but it is unclear whether Gaussian would continue
  to improve and become much better than random search if allowed to
  run for more iterations. Uniform random hill climbing and policy
  gradient descent performed similarly to each other on average, but
  differed greatly on individual runs.}{}

In addition to running strategies to optimize parameterized gaits, we
also wanted to investigate the space of possible gaits.  To accomplish
this, we selected a parameter vector that resulted in motion, but not
an exceptional gait, and plotted performance along each dimension
individually.  In addition, we duplicated each measurement to be able
to estimate the measurement noise at each point.  Results for this
exploration are shown in \figref{explore_dim_1} through
\figref{explore_dim_5}.  As is shown in the figures, some dimensions
are smoother than others, and some measurements are fairly noisy.

\figvarp{explore_dim_1}{.65}{Fitness mean and standard deviation
  vs. dimension 1.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
\figvarp{explore_dim_2}{.65}{Fitness mean and standard deviation
  vs. dimension 2.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
\figvarp{explore_dim_3}{.65}{Fitness mean and standard deviation
  vs. dimension 3.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
\figvarp{explore_dim_4}{.65}{Fitness mean and standard deviation
  vs. dimension 4.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
\figvarp{explore_dim_5}{.65}{Fitness mean and standard deviation
  vs. dimension 5.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}{}

\subsection{Discussion}
% Is your hypothesis supported? What conclusions do the results support
% about the strengths and weaknesses of your method compared to other 
% methods? How can the results be explained in terms of the underlying 
% properties of the algorithm and/or the data.

Because only three trials were tested with each algorithm and no
algorithm consistently outperformed the others, there is a large
standard error for each algorithm, as show in \figref{std_error}. For
this reason, it is unclear if any algorithm is superior to another in
this application. More trials with each algorithm would be necessary
to reach a definitive ranking of the performance of the algorithms.

However, each algorithm discovered at least one gait of over 10 body
lengths/minute, including random. For this reason, we speculate that
the motion model is at least as critical as the learning algorithm
used.

\section{Future work}
% What are the major shortcomings of your current method? For each 
% shortcoming, propose additions or enhancements that would help overcome it.

There are several directions in which we could continue our
work. First, the error margins for our runs were large, so reducing
them by running more trials would lead to more conclusive data. It
would also be insightful to run our algorithms on different motion
models.  We suspect that our choice of motion model influenced the
results greatly, as even random choices in the space produced gaits
that moved a significant fraction of the time (> 5\%).  It would be
interesting to see how learning methods would perform using a model
that included a much higher percentage of unproductive gaits. We also
intend to experiment further with SVM regression and evolutionary
algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
turning, as opposed to it walking long distances. This could be an
interesting learning goal in future projects. To these ends, we
propose the following additions and enhancements:

\begin{itemize}
\item More runs and/or longer runs
\item Different motion representations
\item Better tuning of SVM regression
\item Evolutionary algorithms/HyperNEAT\cite{clune}
\item Learning how to turn
\end{itemize}

\section{Conclusion}
% Briefly summarize the important results and conclusions presented in 
% the paper. What are the most important points illustrated by your 
% work? How will your results improve future research and applications 
% in the area?

We have presented an array of approaches to optimizing a quadrupedal
gait for forward speed.  We have implemented and tested different
learning strategies, including uniform and Gaussian random hill
climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
several new predictive methods based on linear and support vector
regression, and an evolved neural network (HyperNEAT).  We have also
compared these approaches to random search as a baseline. Many of the
methods resulted in walks significantly faster than previously
hand-tuned gaits.

Because only three trials were tested with each algorithm and no
algorithm consistently outperformed the others, there was a large
standard error for each method, as shown in \figref{std_error}. Thus
it was unclear if any algorithm was superior to another in this
application. More trials with each algorithm would be necessary to
reach a definitive ranking. Because each algorithm discovered at least one gait of
over 10 body lengths/minute, including random search, we also
conjectured that the motion representation for the robot is more
integral to forward speed than the learning algorithm.  How to learn
the motion representation, in addition to its parameters, remains an
open problem.

\section{Acknowledgments}
% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
\begin{itemize}
\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
\item Jim T\o rreson, University of Oslo: adviser.
\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
\item Cooper Bills, Cornell University: assisted with Wii tracker development.
\item Anshumali Srivastava, Cornell University: Teaching Assistant.
\end{itemize}

\section{The Team}
% A table listing the names of the people who worked on the project, and who
% did what.

\item Diana Hidalgo: 
\begin{itemize}
\item Simplex (Nelder-Mead)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item Plotting and evaluating results
\item Investigation into using Optitrack IR system for tracking robot
\end{itemize}

\item Sarah Nguyen
\begin{itemize}
\item Random search
\item Uniform random hill climbing
\item Gaussian random hill climbing
\item N-dimensional policy gradient descent
\item Linear regression and prediction
\item \code{RunManager} and \code{Strategy} classes
\item Video editing
\end{itemize}

\item Jason Yosinski
\begin{itemize}
\item SVM regression and prediction
\item Evolutionary Neural Network (HyperNEAT)
\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
\item \code{Parameterized Motion Model} abstract base class
\item \code{SineModel} classes
\item \code{explore\_dimensions}
\item Plotting and evaluating results
\end{itemize}

\section{Appendix}
% Describe the code files you uploaded into CMS

A brief description of the code uploaded to the CMS follows:

\begin{itemize}
\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
  with that strategy.
\item \code{RunManager.py}: Deals with all the details of running the robot,
  including choosing an initial parameter, running the robot multiple times, 
  tracking distance walked, and writing to the log file. Also includes the
  \code{explore\_dimensions} method.
\item \code{Strategy.py}: Contains all the different possible strategies, which
  will be passed as objects in \code{optimize.py}. 
\item \code{Robot.py}: Implements the \code{Robot} class, described in
  \secref{implement}.
\item \code{SineModel.py}: Implements a sine based motion model,
  described in \secref{implement}.
\item \code{Motion.py}: Motion helper functions.
\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
\end{itemize}

% LocalWords:  vertices

>>>>

\section{Experimental Approach}

\subsection{Learning Methods}

\edit{write this.  Old quadratot Method section below}

%\section{Method}
\seclabel{method}

@@

\subsection{Platform details}

\edit{write this.  Old quadratot System Architecture and Implementation section below.}

%\section{System Architecture and Implementation}
\seclabel{implement}

% Describe how you implemented your system and how you structured it. 
% This should give an overview of the system, not a detailed 
% documentation of the code. The documentation of the code is part of 
% the code you hand in. You might want to comment on high-level design 
% decisions that you made. Also explain how you obtained your
% data and any pre-processing you did to it.

@@

\subsection{Fitness evaluation details}

\edit{write this.  Old quadratot Experimental Evaluation section below.}

%\section{Experimental Evaluation}

%\subsection{Methodology}
% What are the criteria you are using to evaluate your method? What
% specific hypotheses does your experiment test? Describe the 
% experimental methodology that you used. What are the dependent and 
% independent variables? For projects in machine learning, what is 
% the training/test data that was used, and why is it realistic or
% interesting? Exactly what performance data did you collect and how 
% are you presenting and analyzing it? Comparisons to competing methods 
% that address the same problem or to variations of your own algorithm 
% are particularly useful.

@@
